Rubin Observatory Pointing Algorithm and Survey Simulation

### 2025-11-10 ms.tex
Updates from Overleaf

There have been many recent developments using Machine Learning algorithms and Large Language Model AIs that have led many to believe a trained AI could be used to schedule observations. Unfortunately, the scale of LSST is so much larger than previous surveys it's not clear there is a data set of adequate size that could be used to train and LSST scheduling AI. There is also growing skeptisism around the feasability of making AIs that do not "hallucinate" \citep{Hicks24}.

@@

Throughout this paper we adopt the LSST nomenclature of a "visit" being potentially mulitple exposures at a single location while each of those exposures is refered to as a "snap". Rubin processing typically combines snaps from a visit before running things like transient detection.

@@

\subsection{Kinematic Model}
The kinematic model computes the time in takes for various motions including
\begin{itemize}
    \item{Azimuthal motions of the dome}
    \item{Altitude motion of the dome mounted wind screen}
    \item{Altitude and azimuthal motions of the telescope}
    \item{Rotational motion of the camera}
    \item{Filter change time}
    \item{telescope settle times}
    \item{filter change}
    \item{shutter motion during a visit}
    \item{open and closed optics loop delays}
\end{itemize}

@@

\begin{figure}
    \epsscale{0.7}
    \plotone{plots/motion_times}
    \epsscale{1}
    \caption{Kinematic motion times for different components.  \label{fig:motion_times}}
\end{figure}

\subsection{Downtimes}

including scheduled and unscheduled downtime---

\subsection{Environmental Models}
Seeing, weather, sky brightness

\begin{figure}
    \centering
    \epsscale{0.7}
    \plotone{plots/seeing_over_time.pdf} \\
    \plottwo{plots/cloudy_over_time.pdf}{plots/cloud_levels.pdf}
    \epsscale{1}
    \caption{Seeing and cloud levels.}
    \label{fig:seeing_clouds}
\end{figure}

\begin{figure}
    \centering
    \epsscale{0.7}
    \plotone{plots/hours_pernight.pdf}
    \epsscale{1}
    \caption{The total amount of possible time per night and the average amount of time
after removing weather downtime. Shaded regions show scheduled downtime. Our modeled
unscheduled downtime is not included in the plot. The year progresses from left to right in
the plot starting on October 1st. The variable length June shutdown starts around Night
250. This general weather and maintenance schedule is the same for all simulations.}
    \label{fig:downtime}
\end{figure}

@@

There is an additional algorithm that controls which filters are loaded at the start of each night. This is configured so that if the moon is 40\% illuminated or more the $grizy$ filters are loaded, otherwise the $ugriy$ filters are loaded.

@@

% Might be enough to discuss here to break into more figures
\begin{figure}
    \epsscale{0.5}
    \plotone{plots/cond_airmass.pdf}\plotone{plots/cond_seeing.pdf} \\
    \plotone{plots/cond_selong.pdf}\plotone{plots/cond_skybright.pdf} \\
    \plotone{plots/cond_slewtime.pdf}
    \epsscale{1}
    \caption{Information generated by the model observatory and passed to the observing decision algorithm. These are Mollweide projections of RA,dec coordinates for October 1, 2023. Areas in gray are unreachable by the telescope because they are low altitude or in the zenith exclusion zone. The sun is setting with an altitude of -17 degrees. All the maps are stored as nside=32 HEALpix arrays. }
    \label{fig:conditions}
\end{figure}

@@

@@

xxx--discuss the general basis functions that typically are used for the MDP surveys.

Most surveys define a fixed set of fields (e.g., DES). Early attempts at scheduling LSST used fixed fields, but this leads to an issue of how to optimally dither pointings to fill chip and raft gaps, as well as distribute field overlap regions.

@@

@@

\subsection{Scheduler Decision Tree}
Describe the basic decision tree that gets built. Tier 1 are ToO observations. Tier 2 are Roman overlap observations. Tier 3 are Deep Drilling Field observations. Tier 4 are observations dedicated to observing time scales of several hours. Tier 5 are taking paired observations with 33 minute time gaps. Tier 6 is similar, but for 15 minute gaps. Tier 7 is earth-interior solar system objects, so sets of 4 observations taken at high airmass in twilight time

@@

>>>>

There have been many recent developments using Machine Learning algorithms and Large Language Model AIs that have led many to believe a trained AI could be used to schedule observations. Unfortunately, the scale of LSST is so much larger than previous surveys it is not clear there is a data set of adequate size that could be used to train and LSST scheduling AI. There is also growing skeptisism around the feasibility of making AIs that do not "hallucinate" \citep{Hicks24}.

@@

Throughout this paper we adopt the LSST nomenclature of a "visit" being potentially multiple exposures at a single location while each of those exposures is refered to as a "snap". Rubin processing typically combines snaps from a visit before running things like transient detection.

The latest simulation \citet{v4.3}.

@@

\subsection{Kinematic Model}
The kinematic model computes the time in takes for various motions including
\begin{itemize}
    \item{Azimuthal motions of the dome}
    \item{Altitude motion of the dome mounted wind screen}
    \item{Altitude and azimuthal motions of the telescope}
    \item{Rotational motion of the camera}
    \item{Filter change time}
    \item{Telescope settle times}
    \item{Filter change}
    \item{Shutter motion during a visit}
    \item{Open and closed optics loop delays}
\end{itemize}

@@

\begin{figure}
    \epsscale{0.7}
    \plotone{plots/motion_times}
    \epsscale{1}
    \caption{Kinematic motion times for different telescope components.  \label{fig:motion_times}}
\end{figure}

\subsection{Down times}

The model observatory includes both scheduled downtime and generates random unscheduled downtime events.  
including scheduled and unscheduled downtime. 
Scheduled downtime events are 7 day stretches for general maintenance and 14 day stretches for mirror recoatings. The total scheduled downtime in the 10-year survey is 140 nights.

Unscheduled downtime events are simulated as 1, 3, 7, or 14 days long. For any given night, there is a 1.37\% chance of a minor event (1 night lost), 0.548\% change of an intermediate event (3 nights down), 0.137\% chance of a major event (7 nights down), and 0.0274\% chance of a catastrophic event (14 nights down). We add additional minor downtime events to year 1, with the probability of a minor event happening within a night starting at 50\% and dropping to $\sim$5\%.

Our default scheduled and unscheduled downtime realization is plotted in Figure~\ref{fig:downtimes}.

\begin{figure}
    \centering
    \epsscale{0.7}
    \plotone{plots/downtimes.pdf}
    \caption{The scheduled and unscheduled downtimes included in default survey simulations. \label{fig:downtimes}}
    \epsscale{1}
\end{figure}

\subsection{Environmental Models}

Seeing, weather, sky brightness

\begin{figure}
    \centering
    \epsscale{0.7}
    \plotone{plots/seeing_over_time.pdf} \\
    \plottwo{plots/cloudy_over_time.pdf}{plots/cloud_levels.pdf}
    \epsscale{1}
    \caption{Seeing and cloud coverage distributions used in survey simulations. }
    \label{fig:seeing_clouds}
\end{figure}

\begin{figure}
    \centering
    \epsscale{0.7}
    \plotone{plots/hours_pernight.pdf}
    \epsscale{1}
    \caption{The total amount of possible time per night and the average amount of time
after removing weather downtime. Shaded regions show scheduled downtime. Our modeled
unscheduled downtime is not included in the plot. The year progresses from left to right in
the plot starting on November 1, 2025. The variable length June shutdown starts around Night
250. This general weather and maintenance schedule is the same for all simulations.}
    \label{fig:downtime}
\end{figure}

@@

There is an additional algorithm that controls which filters are loaded at the start of each night. This is configured so that if the moon is 40\% illuminated or more the $grizy$ filters are available, otherwise the $ugriy$ filters are available.

@@

% Might be enough to discuss here to break into more figures
\begin{figure}
    \epsscale{0.5}
    \plotone{plots/cond_airmass.pdf}\plotone{plots/cond_seeing.pdf} \\
    \plotone{plots/cond_selong.pdf}\plotone{plots/cond_skybright.pdf} \\
    \plotone{plots/cond_slewtime.pdf}
    \epsscale{1}
    \caption{Information generated by the model observatory and passed to the observing decision algorithm. These are Mollweide projections rotated so zenith at the center of the map and east is to the left. Areas in gray are unreachable by the telescope because they are low altitude or in the zenith exclusion zone. The date is set to November 1, 2025 with the sun setting with an altitude of -12 degrees. All the maps are stored as nside=32 HEALpix arrays. }
    \label{fig:conditions}
\end{figure}

@@

\section{Python Objects}
Maybe a section on how the code is organized?

\subsection{Observations}

Two flavors of observation objects, normal and pre-scheduled.

Once an observations is completed, all the relevant metadata of the is filled in.

\subsection{Conditions}

The Conditions object is a data class that holds the current time,

\subsection{Survey}

The Survey class is designed to take as input Conditions objects and information about completed observations. The Survey class computes a reward value that can be used to rank it's desirability compared to other Survey objects, and, when the reward value is finite, returns an array of desired observations.

With this simple API, a Survey object can be constructed with either prescheduled observations, or contain an algorithm for generating observations dynamically.

\subsection{Basis Functions}

\subsection{Features}

@@

xxx--discuss the general basis functions that typically are used for the MDP surveys. The most commonly used basis functions are, 1) slewtime, 2) the 5$\sigma$-limiting depth (which is a function of airmass, seeing, and sky brightness), and 3) the difference between the desired survey footprint and the footprint created by completed observations.

xxx--discussion of how we construct the desired footprint as a function of time. This might need a few figures. Maybe a set showing progression of the footprint through a few years. Then a set showing progression with rolling. Pointing out we have multiple bands so northern observatories should have access to the enhanced alert stream. And that we fiddle the rolling so that there are data releases with fairly uniform coverage.

Most astronomical surveys define a fixed set of fields (e.g., DES xxx-cite). Early attempts at scheduling LSST used fixed fields, but this leads to an issue of how to optimally dither pointings to fill chip and raft gaps, as well as distribute field overlap regions.

@@

xxx--This might be a place to point out that we use truncated float comparisons throughout the code to ensure the results are repeatable cross-platform. Values are typically scaled up by $10^5$\ and then rounded to an integer before comparisons are made. By convention, basis functions and associated weights are kept around order unity.

@@

\subsection{Scheduler Decision Tree}

Describe the basic decision tree that gets built. Tier 1 are ToO observations. Tier 2 are Roman overlap observations. Tier 3 are Deep Drilling Field observations. Tier 4 are observations dedicated to observing time scales of several hours. Tier 5 are taking paired observations with 33 minute time gaps. Tier 6 is similar, but for 15 minute gaps. Tier 7 is earth-interior solar system objects, so sets of 4 observations taken at high airmass in twilight time.

XXX--maybe a table of the different survey tiers, listing total exposure time and filters used for each.

@@

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_DD_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_DD_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_DD_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of deep drilling field and Roman bulge observations. \label{fig:tiers_ddf}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_ToO_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_ToO_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_ToO_and_night_lt_730_HOUR_Hourglass.pdf}
\plotone{plots/tiers/run_name_Hourglass_years_3-5_scheduler_note_like_ToO_and_night_gt_1095_and_night_lt_1825_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of ToO observations. In early years there are fewer ToOs as we assume LIGO starts a run at XXX.\label{fig:tiers_toos}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_blob_long_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_blob_long_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_blob_long_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of initial observations that will later be observed again to sample timescales of several hours. \label{fig:tiers_hours1}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_long_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_long_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_long_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of observations matching those in Figure~\ref{fig:tiers_hours1} \label{fig:tiers_hours2}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_greedy_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_greedy_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_greedy_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of greedy observations. These are last-resort observations when conditions and/or time constraints make all other tiers unviable. \label{fig:tiers_greedy}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_pair_33_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_pair_33_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_pair_33_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of 33 minute paired observations observations .\label{fig:tiers_pair33}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_pair_15_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_pair_15_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_pair_15_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of 15 minute paired observations observations \label{fig:tiers_pair15}}
\end{figure}

\begin{figure}
\epsscale{0.5}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvis_scheduler_note_like_twilight_HEAL_SkyMap.pdf}
\plotone{plots/tiers/baseline_v4_3_1_10yrs_Nvisits_as_function_of_Alt_Az_scheduler_note_like_twilight_HEAL_SkyMap.pdf}
\plotone{plots/tiers/run_name_Hourglass_scheduler_note_like_twilight_and_night_lt_730_HOUR_Hourglass.pdf}
\epsscale{1}
\caption{Distribution of twilight observations optimized for detecting earth-interior solar system objects. \label{fig:tiers_twiinner}}
\end{figure}

\section{Survey Performance}

XXX--Maybe have a section here with a few of the main MAF metrics and show how the baseline survey performs.
